---
title: "Module1 Practice"
author: "Dimitri AGOGUE"
date: "`r Sys.Date()`"
output:
  bookdown::pdf_document2: 
  latex_engine: xelatex
  number_section: False
  
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE)
```

## Loading and cleaning

```{r}
library(tidyverse)
library(DT)
```


### Todo #1 : Help Page for readr::read_cvs

Argument 	Function

- file 		points to the file containing the data
- delim		single character separating fields in a record
- comment	comments identifier (so they can be silently ignored) 
- skip		identifies the number of lines to skip before reading the data.

```{r}
filename <- paste0(
  "https://gedeck.github.io/machine-learning-with-tidymodels/",
  "datasets/penguins_modified.csv.gz"
)

penguins <- readr::read_csv(filename)
```
```{r}
?readr::read_csv
```

```{r}
penguins
```

### Inspect Data to identify missing values

```{r}
colSums(is.na(penguins))
```

```{r}
penguins[rowSums(is.na(penguins))>0, ]
```


```{r}
df <- penguins |> 
  drop_na()

#Alternatively, we could specify columns
df_1 <- penguins |>
  drop_na(bill_length_mm, bill_depth_mm)
```


### Identifying Outliers

```{r}
df <- df |> 
  mutate(id = row_number())
ggplot(df, aes(x = id, y = bill_length_mm))+
  geom_line()
```

```{r}
penguins |>
  mutate(id = row_number(), .before = species)|>
  slice(19:25)|>
  datatable(rownames = FALSE)
```

```{r}
df <- penguins|>
  drop_na()|>
  filter(bill_length_mm < 100)
```


### ToDo #2

**bill_depth_mm**
```{r}
df <- penguins |>
  drop_na()|>
  mutate(id = row_number())
ggplot(df, aes(x = id, y = bill_depth_mm))+
  geom_line()
```

**flipper_length_mm**
```{r}
ggplot(df, aes(x = id, y = flipper_length_mm))+
  geom_line()
```


**flipper_length_mm**
```{r}
ggplot(df, aes(x = id, y = body_mass_g))+
  geom_boxplot()
```

```{r}
penguins|>
  mutate(id = row_number(), .before = species)|>
  filter(body_mass_g > 6000 | body_mass_g < 3000)|>
datatable(rownames = FALSE)
```


## Manipulating data

```{r}
library(nycflights13)
dim(flights)
```

```{r}
df <- flights |>
  # step 1
  filter(distance < 1000) |>
  # step 2
  select(dep_delay, arr_delay, origin, dest, air_time, distance) |>
  # step 3
  mutate(
    Z_dep_delay = (dep_delay - mean(dep_delay, na.rm = TRUE)) /
      sd(dep_delay, na.rm = TRUE),
    dep_delay = dep_delay / 60,
    arr_delay = arr_delay / 60,
    speed = distance / (air_time / 60)
  ) |>
  # step 4
  arrange(-speed) |>
  # step 5
  print(n = 4)
```



## Data Visualization 

```{r}
library(tidyverse)
library(patchwork)
library(GGally)
library(hexbin)
```


```{r}
ggplot(data = mtcars, mapping = aes(x = wt, y = mpg))+
  geom_point(color = "darkgreen")+
  geom_smooth(formula = y ~ x, method = "lm")
```


### Visualizing single variables

```{r}
g1 <- ggplot(data = mtcars, mapping = aes(y = mpg)) +
  geom_boxplot() +
  labs(y = "MPG", title = "Boxplot")
g2 <- ggplot(data = mtcars, mapping = aes(x = mpg)) +
  geom_histogram(bins = 20) +
  labs(x = "MPG", title = "Histogram")
g3 <- ggplot(data = mtcars, mapping = aes(x = mpg)) +
  geom_density() +
  labs(x = "MPG", title = "Density plot")

g1 + g2 + g3 + plot_layout(widths = c(1, 2, 2))
```





```{r}
paste0("graph",".png")
```


:::{.callout-note}

:::









